---
title: "GRTS master sample for MGRS 1 km x 1 km grid covering Belgium and Flanders"
author: "Toon Westra"
output:
  bookdown::html_document2:
    toc: TRUE
    toc_depth: 2
    toc_float:
      collapsed: FALSE
      smooth_scroll: FALSE
    number_sections: TRUE
date: "2023-05-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#remotes::install_github("hrbrmstr/mgrs")
library(tidyverse)
library(sf)
library(giscoR)
library(grtsdb)
library(terra)
library(leaflet)
library(mgrs)
library(conflicted)

conflicts_prefer(dplyr::filter())

# ISO8601 timestamp to set as fixed value in the GeoPackage 
# (to be UPDATED to the actual creation date; at least update for each version):
Sys.setenv(OGR_CURRENT_DATE = "2023-05-03T00:00:00.000Z")
# This is used to keep results reproducible, as the timestamp is otherwise
# updated each time.
# Above environment variable OGR_CURRENT_DATE is used by the GDAL driver.
# The time precision (milliseconds) & timezone (UTC, denoted by 'Z') is
# needed to meet Requirement 15 of the GeoPackage standard (version 1.2.1 & 1.3).
```

# MGRS 1 km grid

## Background

The Military Grid Reference System (MGRS), sometimes called the ‘UTM grid’, is often used to define grid cells with a certain dimension (for example 1 km x 1 km) in biodiversity monitoring. Background information can be found [here](https://inbo.github.io/tutorials/articles/spatial_mgrs).  

## Create MGRS 1 km grid covering Belgium

We create a MGRS 1km grid covering Belgium as follows:

+ load sf object of Belgium using the `giscoR` package (Hernangomez, 2023)
+ transform it to `crs = 25831`
+ take a buffer of 5 km
+ derive the bounding box
+ round the bounding box to a multitude of 1000


```{r}

belgium <- gisco_get_nuts(nuts_level = 2, country = "Belgium", resolution = "01") %>%
  st_transform(crs = 25831)

flanders <- belgium %>%
  filter(NUTS_NAME %in% c("Prov. Antwerpen", 
                          "Prov. Oost-Vlaanderen", 
                          "Prov. West-Vlaanderen", 
                          "Prov. Vlaams-Brabant", 
                          "Prov. Limburg (BE)"))


belgium_buffer <- belgium  %>%
  st_buffer(dist = 5000)

bbox_belgium_buffer <- st_bbox(belgium_buffer)

xmin <- floor(bbox_belgium_buffer$xmin/1000)*1000
ymin <- floor(bbox_belgium_buffer$ymin/1000)*1000
xmax <- ceiling(bbox_belgium_buffer$xmax/1000)*1000
ymax <- ceiling(bbox_belgium_buffer$ymax/1000)*1000

xmin <- xmin - 1000
ymin <- ymin - 1000

x_res <- (xmax - xmin) / 1000
y_res <- (ymax - ymin) / 1000

```



# GRTS sample

We use the `grtsdb` package to create a grts sample with the bounding box created above and a cell size of 1 km.

```{r}

db <- connect_db()

bbox <- rbind(
  x = c(xmin, xmax),
  y = c(ymin, ymax)
)
  
cellsize <- 1000

add_level(bbox = bbox, cellsize = cellsize, grtsdb = db)
 
```
Next we convert the result to a raster, a point shapefile and a polygon shapefile and crop it to the extent of Belgium and Flanders.
We use the R package `mgrs` (Rudis, 2022) to convert xy-coordinates of the SW-corner of a 1km grid cell to the mgrs coordinate at 1 km resolution.

```{r}
sample <- extract_sample(
  grtsdb = db, samplesize = 100000, bbox = bbox, cellsize = cellsize
)

sample_utm <- sample %>%
  rowwise() %>%
  mutate(cellcode = utm_to_mgrs(zone = 31,hemisphere =  "N",easting =   x1c, northing =  x2c, precision = 2))

belgium_buffer_vector_terra <- vect(belgium_buffer)

sample_utm_raster <- sample_utm %>%
  select(x = x1c, y = x2c, ranking) %>%
  rast(type = "xyz", crs = crs(belgium_buffer_vector_terra))

sample_utm_points_sf <- sample_utm %>%
  st_as_sf(coords = c("x1c", "x2c"), crs = 25831)

sample_utm_polygons_sf <- sample_utm %>%
  mutate(geom = str_c("POLYGON ((", x1c - 500, " ", x2c - 500, 
                                ",", x1c + 500, " ", x2c - 500,
                                ",", x1c + 500, " ", x2c + 500,
                                ",", x1c - 500, " ", x2c + 500,
                                ",", x1c - 500, " ", x2c - 500, "))")) %>%
  st_as_sf(wkt = "geom", crs = 25831)

```


```{r}

belgium_buffer_2000 <- belgium %>%
  group_by(CNTR_CODE) %>%
  summarise(level = "Belgium") %>%
  ungroup() %>%
  st_buffer(2000) %>%
  select(level)

grts_utm1_belgium_polygons <- sample_utm_polygons_sf %>%
  st_join(belgium_buffer_2000) %>%
  filter(!is.na(level))

grts_utm1_belgium_points <- sample_utm_points_sf %>%
  filter(cellcode %in% grts_utm1_belgium_polygons$cellcode)

grts_utm1_belgium_raster <- sample_utm_raster %>%
  crop(vect(belgium_buffer_2000)) %>%
  mask(vect(belgium_buffer_2000))

flanders_buffer_2000 <- flanders %>%
  group_by(CNTR_CODE) %>%
  summarise(level = "flanders") %>%
  ungroup() %>%
  st_buffer(2000) %>%
  select(level)

grts_utm1_flanders_polygons <- sample_utm_polygons_sf %>%
  st_join(flanders_buffer_2000) %>%
  filter(!is.na(level))

grts_utm1_flanders_points <- sample_utm_points_sf %>%
  filter(cellcode %in% grts_utm1_flanders_polygons$cellcode)

grts_utm1_flanders_raster <- sample_utm_raster %>%
  crop(vect(flanders_buffer_2000)) %>%
  mask(vect(flanders_buffer_2000))
  
```

# Check result

```{r, eval = FALSE}

# here we check if the results matches with a shapefile used at INBO that contains the mgrs 1 km of Belgium

utm1 <- st_read("../data/utm1.shp") %>%
  st_transform(25831)

utm1_flanders <- utm1 %>%
  st_join(flanders) %>%
  filter(!is.na(NUTS_NAME))

utm1_flanders_centroid <- utm1_flanders %>%
  st_centroid()

check_utm1 <- utm1_flanders %>%
  slice(10000:10100) 

sample_check <- sample_utm_points_sf %>%
  st_join(utm1_flanders) %>%
  filter(!is.na(TAG)) %>%
  mutate(cellcode = str_remove(cellcode, "31U")) %>%
  mutate(check = cellcode == TAG)

check_mgrs_code <- all(sample_check$check)
  
sample_check_show <- sample_check %>%
  filter(cellcode %in% check_utm1$TAG) %>%
  st_transform(4326)

check_utm1 <- check_utm1 %>%
  st_transform(crs = 4326)

check_utm1 %>%
  leaflet() %>%
  addTiles() %>%
  addPolygons() %>%
  addCircleMarkers(data = sample_check_show)
```

## Belgium

The map below show the 50 mgrs cells with the lowest ranking in Belgium.

```{r}
grts_utm1_belgium_polygons_50 <- grts_utm1_belgium_polygons %>%
  slice_min(order_by = ranking, n = 50) %>%
  st_transform(4326)

grts_utm1_belgium_points_50 <- grts_utm1_belgium_points %>%
  filter(ranking %in% grts_utm1_belgium_polygons_50$ranking) %>%
  st_transform(4326)

grts_utm1_belgium_polygons_50 %>%
  leaflet() %>%
  addTiles() %>%
  addPolygons(label = ~ranking, labelOptions = labelOptions(noHide = TRUE)) %>%
  addCircleMarkers(data = grts_utm1_belgium_points_50, color = "red", label = ~ranking)
```

The raster of the grts is shown below. 

```{r}
grts_utm1_belgium_raster %>%
  plot()
```

## Flanders

The map below show the 50 mgrs cells with the lowest ranking in Flanders.

```{r}
grts_utm1_flanders_polygons_50 <- grts_utm1_flanders_polygons %>%
  slice_min(order_by = ranking, n = 50) %>%
  st_transform(4326)

grts_utm1_flanders_points_50 <- grts_utm1_flanders_points %>%
  filter(ranking %in% grts_utm1_flanders_polygons_50$ranking) %>%
  st_transform(4326)

grts_utm1_flanders_polygons_50 %>%
  leaflet() %>%
  addTiles() %>%
  addPolygons(label = ~ranking, labelOptions = labelOptions(noHide = TRUE)) %>%
  addCircleMarkers(data = grts_utm1_flanders_points_50, color = "red", label = ~ranking)
```
The raster of the grts is shown below. 

```{r}
grts_utm1_flanders_raster %>%
  plot()
```


# Output

We write the resulting sf objects as a geopackage file and the raster object as a tif file.

We also save the sqlite file.

```{r}

if (!dir.exists("../output")) {
  
  dir.create("../output")
  
}

st_write(grts_utm1_belgium_points, dsn = "../output/grts_master_utm1_belgium.gpkg",
         layer = "cell_center",
         delete_layer = TRUE)

st_write(grts_utm1_belgium_polygons, dsn = "../output/grts_master_utm1_belgium.gpkg",
         layer = "cell_polygons",
         delete_layer = TRUE)

writeRaster(grts_utm1_belgium_raster, "../output/grts_master_utm1_belgium.tif", overwrite = TRUE)

st_write(grts_utm1_flanders_points, dsn = "../output/grts_master_utm1_flanders.gpkg",
         layer = "cell_center",
         delete_layer = TRUE)

st_write(grts_utm1_flanders_polygons, dsn = "../output/grts_master_utm1_flanders.gpkg",
         layer = "cell_polygons",
         delete_layer = TRUE)

writeRaster(grts_utm1_flanders_raster, "../output/grts_master_utm1_flanders.tif", overwrite = TRUE)


grts_utm1_flanders_polygons %>%
  st_transform(crs = 31370) %>%
  st_write("../output/grts_master_utm1_flanders_lambert72.gpkg", delete_layer = TRUE)

file.copy("grts.sqlite", "../output/grts_master_belgium.sqlite", overwrite = TRUE)
```

# Reference {-}

Hernangomez D (2023). giscoR: Download Map Data from GISCO API - Eurostat. <https://doi.org/10.5281/zenodo.4317946>, <https://ropengov.github.io/giscoR/>

Rudis B (2022). mgrs: Convert 'MGRS' ('Military Grid Reference System') Coordinates From/To Other Coordinate Systems. R package version 0.2.3,
  <https://gitlab.com/hrbrmstr/mgrs>.
